Dark Control:\\ A Unified Account of Default Mode Function\\ by Markov Decision Processes

### 2017-09-17 paper/paper.tex
finished addressing comments

\item Experiment (Animals): We hypothesize that experience replay for browsing problem solutions subserved by the DMN is necessary for choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before make analogous choices in the future. Necessity of the DMN, in addition to the hippocampus, for “mind-searching” actions during choice behavior can be demonstrated by causal dysruption of DMN regions, such as circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.

@@

\subsection{The prefrontal cortex: action consideration and stimulus-value association}
Analogous to the PMC,
the dorsomedial PFC (dmPFC) of the DMN is believed to subserve
multi-sensory processes
across time, space, and content domains to
exert top-level control on behavior.
Comparing to the PMC, however,
dmPFC function may be closer to a
``mental sketchpad'' \citep{goldman1996prefrontal}, as it
potentially subserves the de-novo construction and manipulation
of meaning representations instructed by stored semantics and memories
\citep{bzdok2013segregation}.
The dmPFC may subserve inference, representation, and assessment
of one's own and other individuals' action considerations.
Generally,
neurological patients with tissue damage in the prefrontal cortex
are known to struggle with
adaptation to new stimuli and events
\citep{stuss1986frontal}.
Specifically, neural activity in the human dmPFC
reflected expectations about other peoples' actions and
outcomes of these predictions.
Neural activity in the dmPFC indeed explained the performance decline
of inferring other peoples' thoughts in aging humans \citep{moran2012social}.
Certain dmPFC neurons in macaque monkeys exhibited a preference
for processing others', rather than own, behavior
with fine-grained adjustment of contextual circumstances \citep{yoshida2010neural}.
% In fact, the spatially neighboring dorsal anterior cingulate cortex
% has also been linked to computing values and efforts of
% persisting a behavioral plan versus switching the
% environmental context in several lesion studies \citep{kolling2016value}.
%
Such highly abstract neural computations necessarily rely on the
construction of probabilistic internal information drawing from
memory recall, mental scene elaboration processes,
and stored knowledge of world regularities.
%
Moreover,
in a computational neuroimaging experiment,
dorsomedial PFC activity preferentially processed the consequences of
action choices that were considered but not actually executed,
whereas ventromedial PFC (vmPFC) activity
processed especially value outcomes of actions actually performed on the environment
\citep{nicolle2012agent}.

@@

\section{Reinforcement learning control: a process model for DMN function}
We now argue the outlined neurobiological properties
of the DMN regions
to be sufficient for implementing all components
of a full-fledged RL system.
Recalling past experience, considering candidate actions,
random sampling of possible experiences, as well as
estimation of instantaneous and expected delayed reward outcomes
are key components of intelligent RL agents
that are plausible to functionally intersect in the DMN.

RL is a problem-solving technique in which,
through repeated interactions with an \textit{environment},
an agent learns to reach goals and optimize reward signals
in an iterative trial-and-error fashion (Fig. \ref{fig:rl}).
At a given moment, each taken \textit{action} $a$ triggers a change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a,s')$ collected by the agent.
If the collected reward outcome yields a negative value it can be
more naturally interpreted as \textit{punishment}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction \textemdash or aversion \textemdash accompanying the execution of
a particular action.
% Action may be delayed to achieve substantial increases in expected reward
% that is expected to grow with time.

@@

% \subsection{Learning in the DMN}
\subsubsection{Value approximation and the policy matrix}
As already mentioned in the previous section, Q-learning optimizes over the class of
deterministic policies of the form \eqref{eq:qlearning}. State spaces may be extremely large,
and tracking all possible states and actions may require prohibitively excessive
computation and memory resources.
The need of maintaining an explicit table of
states can be eliminated by instead using of an approximate $Q$-value function $\tilde{Q}(s,a|\theta)$
by keeping track of an approximating parameter $\theta$ of much lower dimension than the number of states.
At a given time step, the world is in a state $s \in \mathcal S$, and the agent takes an
action which it expects to be the most valuable on average, namely

@@

\section{Conclusion}
Which brain function could be important enough
for the existence and survival of the human species
to justify constantly high energy costs?
MDPs motivate an attractive
formal account of how the human association cortex
might implement multi-sensory representation and control of the environment to
optimize the organism's interaction with the world.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by simple but non-trivial mechanisms.
%
From the view of a Markovian sequential decision process,
human behavior unfolds by integrating action outcomes from stored past events
and extrapolation to
upcoming events for guiding action choice in the present context.
MDPs also provide a mathematical formalism how
opportunity in the environment can be recursively exploited
when confronted with challenging decisions.
%
This functional interpretation may well be compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world representation
by injecting stochasticity into the recall of past
actions and the estimation of action outcomes may explain why
highly accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.

It is an important feature of the proposed artificial intelligence perspective
on DMN biology that it is practically computable and
yields falsifiable neuroscientific hypotheses.
%
Neuroscience experiments could be designed that operationalize
the set of action, value, and state variables that govern
the behavior of intelligent RL agents.
%
At the least, we propose an alternative vocabulary to
describe, contextualize, and interpret experimental findings in neuroscience studies
on the DMN.
%
Ultimately,
DMN activity may instantiate a holistic integration
ranging from real experience over purposeful dreams to anticipated futures
for continued refinement of the organism's fate.

>>>>

\item Experiment (Animals): We hypothesize that experience replay for browsing problem solutions subserved by the DMN is necessary for choice behavior in mice. Hippocampal single-cell recordings have shown that neural patterns during experimental choice behavior are reiterated during sleep and before make analogous choices in the future. Necessity of the DMN, in addition to the hippocampus, for “mind-searching” actions during choice behavior can be demonstrated by causal disruption of DMN regions, such as circumscribed brain lesion or optogenetic intervention in the inferior parietal and prefrontal cortices.

@@

\subsection{The prefrontal cortex: action consideration and stimulus-value association}
Analogous to the PMC,
the dorsomedial PFC (dmPFC) of the DMN is believed to subserve
multi-sensory processes
across time, space, and content domains to
exert top-level control on behavior.
Comparing to the PMC, however,
dmPFC function may be closer to a
``mental sketchpad'' \citep{goldman1996prefrontal}, as it
potentially subserves the de-novo construction and manipulation
of meaning representations instructed by stored semantics and memories
\citep{bzdok2013segregation}.
The dmPFC may subserve inference, representation, and assessment
of one's own and other individuals' action considerations.
Generally,
neurological patients with tissue damage in the prefrontal cortex
are known to struggle with
adaptation to new stimuli and events
\citep{stuss1986frontal}.
Specifically, neural activity in the human dmPFC
reflected expectations about other peoples' actions and
outcomes of these predictions.
Neural activity in the dmPFC indeed explained the performance decline
of inferring other peoples' thoughts in aging humans \citep{moran2012social}.
Certain dmPFC neurons in macaque monkeys exhibited a preference
for processing others', rather than own, behavior
with fine-grained adjustment of contextual circumstances \citep{yoshida2010neural}.
% In fact, the spatially neighboring dorsal anterior cingulate cortex
% has also been linked to computing values and efforts of
% persisting a behavioral plan versus switching the
% environmental context in several lesion studies \citep{kolling2016value}.
%
% Such highly abstract neural computations necessarily rely on the
% construction of probabilistic internal information drawing from
% memory recall, mental scene elaboration processes,
% and stored knowledge of world regularities.
% %
% Moreover,
% in a computational neuroimaging experiment,
% dorsomedial PFC activity preferentially processed the consequences of
% action choices that were considered but not actually executed,
% whereas ventromedial PFC (vmPFC) activity
% processed especially value outcomes of actions actually performed on the environment
% \citep{nicolle2012agent}.

@@

\section{Reinforcement learning control: a process model for DMN function}
We now argue the outlined neurobiological properties
of the DMN regions
to be sufficient for implementing all components
of a full-fledged reinforcement-learning (RL) system.
Recalling past experience, considering candidate actions,
random sampling of possible experiences, as well as
estimation of instantaneous and expected delayed reward outcomes
are key components of intelligent RL agents
that are plausible to functionally intersect in the DMN.

RL is an area of machine learning concerned with learning optimal behavioral through interactions
with an \textit{environment}, the goal being to maximize some notion of \textit{cumulative reward}.
The optimal behavior typical
takes the future into account, as some rewards could be \textit{delayed}.
Through repeated interactions with the environment,
the agent learns to reach goals and optimize reward signals
in an iterative trial-and-error fashion (Fig. \ref{fig:rl}).
At a given moment, each taken \textit{action} $a$ triggers a change
in the \textit{state} of the environment
$s \rightarrow s'$, accompanied by environmental feedback signals as \textit{reward}
$r = r(s, a,s')$ collected by the agent.
If the collected reward outcome yields a negative value it can be
more naturally interpreted as \textit{punishment}.
In this view, the environment is partially controlled by
the action of the agent and the reward can be thought
of as satisfaction \textemdash or aversion \textemdash accompanying the execution of
a particular action.
% Action may be delayed to achieve substantial increases in expected reward
% that is expected to grow with time.

@@

% \subsection{Learning in the DMN}
\subsubsection{Value approximation and the policy matrix}
\label{sec:policymat}
As already mentioned in the previous section, Q-learning optimizes over the class of
deterministic policies of the form \eqref{eq:qlearning}. State spaces may be extremely large,
and tracking all possible states and actions may require prohibitively excessive
computation and memory resources.
The need of maintaining an explicit table of
states can be eliminated by instead using of an approximate $Q$-value function $\tilde{Q}(s,a|\theta)$
by keeping track of an approximating parameter $\theta$ of much lower dimension than the number of states.
At a given time step, the world is in a state $s \in \mathcal S$, and the agent takes an
action which it expects to be the most valuable on average, namely

@@

\section{Conclusion}
Which brain function could be important enough
for the existence and survival of the human species
to justify constantly high energy costs?
MDPs motivate an attractive
formal account of how the human association cortex
might implement multi-sensory representation and control of the environment to
optimize the organism's interaction with the world.
% by obtaining reward and avoiding pain.
This idealized process model explains
a number of previous experimental observations in the
DMN by simple but non-trivial mechanisms.
%
From the view of a Markovian sequential decision process,
human behavior unfolds by integrating action outcomes from policy matrix (mapping from states to actions which maximize expected cumulative reward; see subsection \ref{sec:policymat})
% stored past events
and extrapolation to
upcoming events for guiding action choice in the present context.
MDPs also provide a mathematical formalism how
opportunity in the environment can be recursively exploited
when confronted with challenging decisions.
%
This functional interpretation may well be compatible with the DMN's
poorly understood involvement across
autobiographical memory recall, problem solving,
abstract reasoning, social cognition,
as well as delay discounting and self-related prospection.
Improvement of the internal world representation
by injecting stochasticity into the recall of past
actions and the estimation of action outcomes may explain why
highly accurate memories have been disfavored in human evolution
and why human creativity might be adaptive.

It is an important feature of the proposed artificial intelligence perspective
on DMN biology that it is practically computable and
yields falsifiable neuroscientific hypotheses.
%
Neuroscience experiments could be designed that operationalize
the set of action, value, and state variables that govern
the behavior of intelligent RL agents.
%
At the least, we propose an alternative vocabulary to
describe, contextualize, and interpret experimental findings in neuroscience studies
on the DMN.
%
Ultimately,
DMN activity may instantiate a holistic integration
ranging from real experience over purposeful dreams to anticipated futures
for continued refinement of the organism's fate. Indeed, a major hurdle in explaining DMN activity from brain-imaging studies has been its very similar engaged across time scales: thinking about the past (e.g., autobiographical memory), thinking about hypothetical presents (e.g., dreams during sleep or daytime mind-wandering), and thinking about future scenarios (e.g., prospection or delay discounting). The MDP account of DMN function naturally integrates this a-priori diverging mental processes into a common framework. % It is this feature of our functional view that we underline in the referred sentence.
